Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-04-06 paper/paper.tex
Capitalization of titles Cf. https://en.wikipedia.org/wiki/Colon_(punctuation)#Use_of_capitals

\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_overview_DMN.pdf}
  \caption{Neurobiological overview of the DMN ... \textbf{Please add a figure caption}}
  \label{fig:neurobiological_overview_DMN}
\end{figure}
%
\subsection{The posteromedial cortex: Global monitoring and information integration}
\suggestremove{Among all parts of the human brain,}
the midline structures of the DMN,
including the posteromedial cortex (PMC) and
the medial prefrontal cortex (mPFC),
are responsible for the highest metabolic turn-over
of glucose energy consumption \citep{raichle2001pnas}.
These metabolic characteristics go hand-in-hand with
brain-imaging analyses \citep{andrews2010} in
suggesting that the PMC and \suggestremove{d}mPFC
potentially represent the functional backbone of the DMN.

@@

\subsection{The prefrontal cortex: Stimulus-value association and action consideration}
Analogous to the PMC,
the dorsomedial PFC (dmPFC) is believed to subserve predominantly ambiguous processes
across time, space, and content domains in
sensory-independent top-down pathways.
In contrast to the PMC, however,
the dmPFC may be the DMN node whose function comes closest to a
``mental sketchpad'' \citep{goldman1996prefrontal}, as it
potentially subserves the de-novo generation and binding
of meaning representations instructed by stored semantics and memories.

@@

\subsection{The hippocampus: Memory, space, and experience replay}
The midline core of the DMN is probably in constant communication
with the HC in the medial temporal lobe \citep{vincet2006}.
The HC is long known to be involved in memory and spatial navigation in animals and humans.

@@

\subsection{The right and left TPJ: Prediction error signaling and world semantics}
While the DMN emerges with its midline structures early in human development,
the right and left TPJ may become fully integrated into this canonical
network only after birth \citep{doria2010}.
The TPJs are known to have hemispheric differences
according to their cytoarchitectonic borders and gyrification pattern
\citep{seghier2013angular}.
Neuroscientific investigations on hemispheric functional specialization
have highlighted the right versus left cerebral hemisphere as dominant for
attention versus language functions.

@@

\section{Reinforcement learning: A process model for DMN function}
The outlined neurobiological properties and suspected functions of the DMN
are argued to be sufficient
for implementing all components
of a full-fledged \textit{reinforcement learning (RL)} algorithm.
Integration of past experience, probabilistic mapping of candidate actions,
random sampling of possible scenarios as well as
incorporation of instantaneous, delayed, and expected reward outcomes
are key components of intelligent RL agents
that are plausible to cross-talk in the DMN.

@@

\subsubsection{Excursion: The role of the hippocampus}
\paragraph{Does the hippocampus do Monte-Carlo sampling?}
In RL, Monte-Carlo simulation can be used to update the agent's belief state
\citep{silver2010monte}. Such methods have a sample complexity that is determined only by the underlying difficulty of the POMDP, rather than the size of the state space or observation space,
which can be prohibitively large.
Monte-Carlo simulation provides a simple method for evaluating the value / fitness of a state.
They provide an effective mechanism both for tree search and for belief state updates, breaking the curse of dimensionality and allowing much greater scalability than a RL agent without stochastic resampling procedures.

@@

\subsection{Predictive Coding Hypothesis}
The predictive coding framework
\citep{clark2013whatever, friston2008hierarchical}
is a frequently evoked idea related to the default mode function
\citep{bar2007}.
According to this cognitive account,
cortical responses emerge from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
This dynamic intercourse enables inference on the world by reconciling
gaps between fresh sensory input and stored prior information.
Feed-forward sensory experience is constantly calibrated by
top-down modulation at various hierarchical processing levels.
In fact, axonal
back projections probably outnumber by far the input projection
existing in the brain \citep{salin1995corticocortical}.
This can explain the constructive, generative nature of sensory perception
\citep{friston2010free} and
why motor action is intimately related to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
At each stage of neural processing,
an internally generated prediction of the sensory input is
directly compared against the actual input from the environment.
A prediction error at one of the processing levels,
i.e. the difference between sensory input
and internally predicted sensation,
incurs plasticity changes of neuronal back projections (i.e., model parameters)
for gradually improved future prediction of the environment.
Contextual integration is hence maintained by top-down modulation of sensorimotor
processing by a-priori information about the environment.
The generative model of how perceived sensory events arise in the
world would be incorporated into
the current neuronal wiring.
Indeed,
This process permits updates of the internal model of the environment
to best fit the constant influx of sensory examples.

@@

\subsection{Semantic Hypothesis}
Another frequently proposed cognitive account of DMN function revolves
around forming logical associations and analogies between
the current experience and
the conceptual knowledge derived from past experiences
\citep{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge (i.e., semantics) extractable from the environment
\citep{bar2009proactive}.
In this way, the encoding of complex environments could be facilitated
by relative similarity associations between states.
%
Going much beyond human language ifself,
semantic building blocks provide the basis to
mentally envision non-existent scenarios
that would improve optimal behavior in the environment
by simulating future events.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories
\citep{hassabis2009construction}.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and are associated with
concurrent deficits in forming and integrating relationships between
items \citep{addis2008age, spreng2006temporal}.
Further,
a constructive account explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
Cognitive aspects of egocentric-allocentric switching
are also closely related to episodic memory, language, problem solving,
planning, estimating other people's thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\citep{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \citep{tolman1948cognitive}.

@@

\subsection{Sentinel Hypothesis}
Processing self-relevant information was perhaps the first
cognitive account that was proposed for the DMN \citep{gusnard2001medial}.
Since then,
many investigators have speculated that neural activity in the DMN
may reflect the brain's continuous tracking of
relevance in the environment
as an advantageous evolutionary adaptation.
According to this cognitive account, the human brain's baseline realizes
a ``radar'' function to
detect subjectively salient objects and unexpected events in the environment.
Propositions of such a sentinel function have however seldom detailed
the mechanisms of
how attention and memory resources are exactly reallocated when
encountering a salient environmental stimulus.
In our formal account,
cascades of candidate actions
are recursively explored by human agents. Conversely,
certain branches of candidate action trajectories
are detected to be less worthy to get mentally explored.
This mechanism, expressed by the Bellman equation,
directly implies stratifications over attention and working memory
to relevant objects and events in the environment.
%
Further,
our account provides an explanation for the known implication
of the DMN
in certain goal-directed experimental tasks and
in the idlying task-unconstrained cognitive sets.
Both cognitive sets may entail DMN recruitment
if environmental relevance is processed, manipulated, and used for prediction.
During tasks,
the policy and value estimates are updated to optimize short-term action.
At rest, these parameter updates may
improve especially mid- and long-term action.
We thus provide answers to the unsettled question why the involvement
of the same neurobiological brain circuit, the DMN, has been documented
for baseline house-keeping functions and specific task performances.

@@

\paragraph{Solution: The reparametrization trick.}
\begin{itemize}
\item Choose $\epsilon \sim p_{\text{noise}}$ (noise distribution, independent of $\W^R$!)
\item Set $\z = g(\W^{R}, \x, \epsilon)$, where $g$ is an appropriate class $\mathcal C^1$ function
  \begin{itemize}
  \item results in a sample $\z \sim p_R(.|\x)$, from the correct
    posterior
    \end{itemize}
\end{itemize}
The mapping $g$ should be taught of as a ``blurring'' function which produces noisy versions $\z$,
called \textit{sensations}, of the true world state $\x$.
The result is a scheme for training DBNs via good-old backprop!
Refer to Fig. vae.pdf.
% \ref{fig:vae}.
Some examples of the reparametrization trick for a number of
choices of the posterior distribution are given in Tab. \ref{tab:rptrick}.

>>>>

\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{neurobiological_overview_DMN.pdf}
  \caption{Neurobiological overview of the DMN ... \textbf{Please add a figure caption}}
  \label{fig:neurobiological_overview_DMN}
\end{figure}
%
\subsection{The posteromedial cortex: global monitoring and information integration}
\suggestremove{Among all parts of the human brain,}
the midline structures of the DMN,
including the posteromedial cortex (PMC) and
the medial prefrontal cortex (mPFC),
are responsible for the highest metabolic turn-over
of glucose energy consumption \citep{raichle2001pnas}.
These metabolic characteristics go hand-in-hand with
brain-imaging analyses \citep{andrews2010} in
suggesting that the PMC and \suggestremove{d}mPFC
potentially represent the functional backbone of the DMN.

@@

\subsection{The prefrontal cortex: stimulus-value association and action consideration}
Analogous to the PMC,
the dorsomedial PFC (dmPFC) is believed to subserve predominantly ambiguous processes
across time, space, and content domains in
sensory-independent top-down pathways.
In contrast to the PMC, however,
the dmPFC may be the DMN node whose function comes closest to a
``mental sketchpad'' \citep{goldman1996prefrontal}, as it
potentially subserves the de-novo generation and binding
of meaning representations instructed by stored semantics and memories.

@@

\subsection{The hippocampus: memory, space, and experience replay}
The midline core of the DMN is probably in constant communication
with the HC in the medial temporal lobe \citep{vincet2006}.
The HC is long known to be involved in memory and spatial navigation in animals and humans.

@@

\subsection{The right and left TPJ: prediction error signaling and world semantics}
While the DMN emerges with its midline structures early in human development,
the right and left TPJ may become fully integrated into this canonical
network only after birth \citep{doria2010}.
The TPJs are known to have hemispheric differences
according to their cytoarchitectonic borders and gyrification pattern
\citep{seghier2013angular}.
Neuroscientific investigations on hemispheric functional specialization
have highlighted the right versus left cerebral hemisphere as dominant for
attention versus language functions.

@@

\section{Reinforcement learning: a process model for DMN function}
The outlined neurobiological properties and suspected functions of the DMN
are argued to be sufficient
for implementing all components
of a full-fledged \textit{reinforcement learning (RL)} algorithm.
Integration of past experience, probabilistic mapping of candidate actions,
random sampling of possible scenarios as well as
incorporation of instantaneous, delayed, and expected reward outcomes
are key components of intelligent RL agents
that are plausible to cross-talk in the DMN.

@@

\subsubsection{Excursion: the role of the hippocampus}
\paragraph{Does the hippocampus do Monte-Carlo sampling?}
In RL, Monte-Carlo simulation can be used to update the agent's belief state
\citep{silver2010monte}. Such methods have a sample complexity that is determined only by the underlying difficulty of the POMDP, rather than the size of the state space or observation space,
which can be prohibitively large.
Monte-Carlo simulation provides a simple method for evaluating the value / fitness of a state.
They provide an effective mechanism both for tree search and for belief state updates, breaking the curse of dimensionality and allowing much greater scalability than a RL agent without stochastic resampling procedures.

@@

\subsection{Predictive coding hypothesis}
The predictive coding framework
\citep{clark2013whatever, friston2008hierarchical}
is a frequently evoked idea related to the default mode function
\citep{bar2007}.
According to this cognitive account,
cortical responses emerge from continuous functional interaction between
higher and lower levels of the neural processing hierarchy.
This dynamic intercourse enables inference on the world by reconciling
gaps between fresh sensory input and stored prior information.
Feed-forward sensory experience is constantly calibrated by
top-down modulation at various hierarchical processing levels.
In fact, axonal
back projections probably outnumber by far the input projection
existing in the brain \citep{salin1995corticocortical}.
This can explain the constructive, generative nature of sensory perception
\citep{friston2010free} and
why motor action is intimately related to sensory expectations
\citep{wolpert1995internal, kording2004bayesian}.
At each stage of neural processing,
an internally generated prediction of the sensory input is
directly compared against the actual input from the environment.
A prediction error at one of the processing levels,
i.e. the difference between sensory input
and internally predicted sensation,
incurs plasticity changes of neuronal back projections (i.e., model parameters)
for gradually improved future prediction of the environment.
Contextual integration is hence maintained by top-down modulation of sensorimotor
processing by a-priori information about the environment.
The generative model of how perceived sensory events arise in the
world would be incorporated into
the current neuronal wiring.
Indeed,
This process permits updates of the internal model of the environment
to best fit the constant influx of sensory examples.

@@

\subsection{Semantic hypothesis}
Another frequently proposed cognitive account of DMN function revolves
around forming logical associations and analogies between
the current experience and
the conceptual knowledge derived from past experiences
\citep{bar2007proactive, binder1999conceptual}.
Analogies might naturally tie incoming novel sensory stimuli to
explicit world knowledge (i.e., semantics) extractable from the environment
\citep{bar2009proactive}.
In this way, the encoding of complex environments could be facilitated
by relative similarity associations between states.
%
Going much beyond human language ifself,
semantic building blocks provide the basis to
mentally envision non-existent scenarios
that would improve optimal behavior in the environment
by simulating future events.
Such cognitive processes can afford
the internal generation of necessary information
that is not presented in the surrounding environment
by recombining building blocks of
concept knowledge and episodic memories
\citep{hassabis2009construction}.
Indeed, in aging humans, remembering the past and imaging the future
equally decreased in the level of detail and are associated with
concurrent deficits in forming and integrating relationships between
items \citep{addis2008age, spreng2006temporal}.
Further,
a constructive account explains the reciprocal relationship
between an egocentric first person perspective and
an allocentric bird’s eye perspective immersed in
self-reflection, social knowledge, and autobiographical memories.
%
Cognitive aspects of egocentric-allocentric switching
are also closely related to episodic memory, language, problem solving,
planning, estimating other people's thoughts, and spatial navigation
as these necessitate abstract world knowledge and abstract assocations
for binding the constituent elements in mental scene construction
\citep{schacter2007remembering}.
These scene generation processes could contribute to interpreting the
present and foretelling the future.
This view is for instance supported by evidence in animals that
could learn a \textit{cognitive map of the environment},
even without reward incentives, and exploit it later
for other means \citep{tolman1948cognitive}.

@@

\subsection{Sentinel hypothesis}
Processing self-relevant information was perhaps the first
cognitive account that was proposed for the DMN \citep{gusnard2001medial}.
Since then,
many investigators have speculated that neural activity in the DMN
may reflect the brain's continuous tracking of
relevance in the environment
as an advantageous evolutionary adaptation.
According to this cognitive account, the human brain's baseline realizes
a ``radar'' function to
detect subjectively salient objects and unexpected events in the environment.
Propositions of such a sentinel function have however seldom detailed
the mechanisms of
how attention and memory resources are exactly reallocated when
encountering a salient environmental stimulus.
In our formal account,
cascades of candidate actions
are recursively explored by human agents. Conversely,
certain branches of candidate action trajectories
are detected to be less worthy to get mentally explored.
This mechanism, expressed by the Bellman equation,
directly implies stratifications over attention and working memory
to relevant objects and events in the environment.
%
Further,
our account provides an explanation for the known implication
of the DMN
in certain goal-directed experimental tasks and
in the idlying task-unconstrained cognitive sets.
Both cognitive sets may entail DMN recruitment
if environmental relevance is processed, manipulated, and used for prediction.
During tasks,
the policy and value estimates are updated to optimize short-term action.
At rest, these parameter updates may
improve especially mid- and long-term action.
We thus provide answers to the unsettled question why the involvement
of the same neurobiological brain circuit, the DMN, has been documented
for baseline house-keeping functions and specific task performances.

@@

\paragraph{Solution: the reparametrization trick.}
\begin{itemize}
\item Choose $\epsilon \sim p_{\text{noise}}$ (noise distribution, independent of $\W^R$!)
\item Set $\z = g(\W^{R}, \x, \epsilon)$, where $g$ is an appropriate class $\mathcal C^1$ function
  \begin{itemize}
  \item results in a sample $\z \sim p_R(.|\x)$, from the correct
    posterior
    \end{itemize}
\end{itemize}
The mapping $g$ should be taught of as a ``blurring'' function which produces noisy versions $\z$,
called \textit{sensations}, of the true world state $\x$.
The result is a scheme for training DBNs via good-old backprop!
Refer to Fig. vae.pdf.
% \ref{fig:vae}.
Some examples of the reparametrization trick for a number of
choices of the posterior distribution are given in Tab. \ref{tab:rptrick}.
